Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-15 latex/main.tex
Updates from Overleaf

\subsubsection*{Identification of robust recombination events}

[TODO This section needs a full pass. I've merged in a section that was in the
supplement with an extremely terse section on counting supporting loci.
Reference Figure~\ref{fig:recombinant_qc} for the scatter plot and 
labelled quadrants, and Figure~\ref{fig:copying_pattern_examples}
for examples of the corresponding copying patterns. Also reference
any additional examples that we include beyond A-D.]

% This para was *{Counting supporting loci for recombinant parents}

When counting such loci, we considered nearly adjacent sites (separated by at
most three bases) as single loci, to avoid over-counting support from
correlated loci, which would lead to artificial signals of recombination. Also,
when counting supporting loci for a focal parent, we penalized loci where the
character(s) in the recombinant matches that in the non-focal parent but not in
the focal parent.

Before characterizing recombination events in detail, we sought to remove
probable artifacts. Systematic errors in SARS-CoV-2 sequences associated with
sequencing protocols have been documented (Turakhia et al., 2022), which can
confound recombination detection (De Maio et al., 2024). By inspecting copying
patterns at sites that differ between recombinant parents, we find a convincing
signal when treating adjacent and near-adjacent sites as a single locus.
Artifactual recombination events can be characterized as having fewer than four
such loci supporting the left or right parent on either side of a suggested
breakpoint (STAR Methods). Of the 929 recombination events in the ARG, this
measure identifies 543 as artifactual, leaving 386 ``robust'' events (Figure
S1A). The copying patterns depicted in Figure S1B illustrate the genomic
evidence for three recombination events identified as artifactual, compared to
a marginally robust one.

Evidence that our simple filter effectively flags artifacts can be seen in the
clear association between the AmpliSeq V1 kit and recombination events that we
identify as artifactual. This primer scheme was used to sequence XX of the
samples associated with an artifactual recombination event, in contrast to 15
of the ``robust'' events (Figure 2A), strongly implicating it as a common source
of problematic data. Through further inspection, we observed that many
artifactual recombination events contain stretches of missing data
characteristic of amplicon dropouts (REF).

>>>>

\subsubsection*{Quality control of recombinants}

[TODO This section needs a full pass. I've merged in a section that was in the
supplement with an extremely terse section on counting supporting loci.
Reference Figure~\ref{fig:recombinant_qc} for the scatter plot and 
labeled quadrants, and Figure~\ref{fig:copying_pattern_examples}
for examples of the corresponding copying patterns. Also reference
any additional examples that we include beyond A-D.]

Before characterizing recombination events in detail,
we sought to remove probable artifacts.
Systematic errors in SARS-CoV-2 sequences
associated with sequencing protocols \cite{Turakhia2020}
can complicate recombination detection \cite{DeMaio2024}.
% Explain to readers what convincing genomic evidence looks like.
By inspecting copying patterns at sites that differ between recombinant parents,
we find a convincing signal when treating adjacent and near-adjacent sites as a single locus.
Artifactual recombination events can be characterized
as having fewer than four such loci supporting the left or right parent
on either side of a suggested breakpoint.
When counting such loci, we considered nearly adjacent sites
(separated by at most three bases) as single loci,
to avoid over-counting support from correlated loci,
which could lead to artificial signals of recombination.
Also, when counting supporting loci for a focal parent,
we penalized loci where the character(s)
in the recombinant matches that in the non-focal parent
but not in the focal parent.

Of the 929 recombination events in the ARG,
this measure identifies 543 as artifactual,
leaving 386 ``robust'' events (Figure~\ref{fig:recombinant_qc}).
The copying patterns depicted Figure~\ref{fig:copying_pattern_examples}
illustrate the genomic evidence for three recombination events identified as artifactual,
compared to a marginally robust one.

Evidence that our simple filter effectively flags artifacts can be seen in the
clear association between the AmpliSeq V1 kit and recombination events that we
identify as artifactual. This primer scheme was used to sequence XX of the
samples associated with an artifactual recombination event, in contrast to 15
of the ``robust'' events (Figure 2A), strongly implicating it as a common source
of problematic data.
